The corona of a fully convective star with a near-polar flare

### 2023-12-19 src/tex/ms.tex
[showyourwork] overleaf sync

@@

For the X-ray flare in Fig.~\ref{fig:lightcurves}, we obtain a total energy of \eepic in the $0.2-2\,$keV band, using $L_{X,\rm quiescent}=$\LXquiet\unskip, $L_{X,\rm flaring}=$\LXflaring\unskip, and $\Delta t =$\xraydur\unskip (see Fig.~\ref{fig:lightcurves}) in Eq.~\ref{eq:xrayflare}. The energy of the corresponding OM flare is \eom\unskip. We use the 36ks observing baseline of \textit{XMM-Newton} to calculate a flare rate in OM. Fig.~\ref{fig:ffd} illustrates that it is roughly consistent, if slightly higher, than the extension of \textit{TESS}' flare frequency distributions to lower energies. About one flare per day of the order of the observed one or above can therefore be expected in future X-ray observations of TIC~277. However, this assumes that all optical flares above this energy in the optical data correspond to detected X-ray flares in the EPIC instruments, which is not always the case~\citep{paudel2021simultaneous}. \citet{guarcello2019simultaneous, kuznetsov2021stellar} find moderate correlation between flare energies in Kepler/K2 data and \textit{XMM-Newton}, but the majority of their sampled flare energies are two orders of magnitude above the flare in this work, both in X-ray and optical. \citet{kuznetsov2021stellar} compare primary Kepler mission observations with \textit{XMM-Newton} to find a similar result. We note that the X-ray flare light curve follows the typical fast-rise exponential decay shape, and is therefore unlikely to be eclipsed by the star~\citep{johnstone2012soft}.

\begin{figure}
    \script{FIGURE_tess_ffd.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/277539431_tess_ffd.png}
        \caption{
         Cumulative flare frequency distribution of \textit{TESS} flares (black dots), with a power law fit (green line), and uncertainties (dotted lines). The rate of OM flares (gray square) is calculated from the observing baseline of \textit{XMM-Newton}. 
        }
        \label{fig:ffd}
    \end{centering}
\end{figure}

@@

Bearing in mind the relatively low two-temperature resolution of TIC~277's X-ray spectrum and the dearth of X-ray spectra for stars at the bottom of the main sequence, it shows a coronal temperature make-up common for mid-to-late M dwarf in the saturated activity regime that it will possibly keep for gigayears until it spins down to very low rotation rates~\citep{medina2022galactic, engle2023living}.

@@

At later spectral type, a decline in coronal activity is expected as the magnetosphere transitions from a stellar to a planetary one~\citep{pineda2017panchromatic}. \citet{stelzer2022first} investigated late M and L dwarfs with eROSITA to show that $L_X/L_{\rm bol}$ increases towards later spectral types~\citep{magaudda2022firsta}, but begins to decrease again from M7 downward. In their compilation, together with previous work from \citet{stelzer2012ultracool, cook2014trends, deluca2020extras, williams2014trends, berger2008simultaneous}, TIC~277 appears at the lower end\textcolor{red}{, but consistent with} the $L_X/L_{\rm bol}$ distribution for M7 dwarfs. However, the rotation periods are unknown for most of these late M dwarfs. Sometimes projected rotation periods are known~\citep{cook2014trends}, but true periods are only detected in a handful of these stars, which we list in Table~\ref{tab:latmlxlbol}. TIC~277 appears \textcolor{red}{consistent with} this small sample.

\begin{table*}[]
    \centering
    \caption{Late M dwarfs with known rotation period and X-ray luminosity.}
    \begin{tabular}{lcccc}
        ID & SpT &$P_{\rm rot}$ [d] & $\log_{10} \frac{L_X}{L_{\rm bol}} $ & energy range [keV]\\\hline
        TIC 277 (this work) & M7 & 0.19 & $-4$ & $0.2-2$ \\
        TRAPPIST-1 [1] & M8 & 3.3 & $-4$ & $0.3-10$\\
        LHS 248 [2] & M6.5 & 0.46 & $-3.9$ & $0.2-2$\\
        NLTT 33370 AB [3] & M7 & 0.16 & $-3.5$ & $0.2-2$ \\
        LP 412-31  [4] & M8 & 0.61 & $-3.1$& \textcolor{red}{soft X-ray band?}\\\hline
    \end{tabular}
     \footnotesize
     \vspace{0.1cm}\newline
    [1] \citet{brown2023coronal}, 
    [2] \citet{cook2014trends}, 
    [3] \citet{williams2015simultaneous}, 
    [4] \citet{stelzer2006simultaneous}    
    %3.3 d from \citep{luger2017sevenplanet}
    \label{tab:latmlxlbol}
\end{table*}

@@

\section{Summary and Conclusions}
\label{sec:summary}
We investigated whether TIC 277, a rapidly rotating M7 dwarf with a flare localized at $81^{\circ}$ latitude, shows unusual coronal or flaring properties that could explain the occurrence of this flare so close to the rotational pole. We obtained 36 ks of EPIC and OM observations with \textit{XMM-Newton}, and studied the five Sectors of red-optical 27-day light curves provided by \textit{TESS}. We found a mean quiescent coronal temperature of \Tqmean, and a flaring rate $f(> \log_{10} E_{\rm flare, bol} = 31.5\,\rm{erg})\approx$\ffdr with energy distribution with a slope of \ffdalpha, all typical of fully convective M dwarfs in the saturated regime. We also detected a simultaneous X-ray and optical flare with an energy of \eepic in the $0.2-2\,$keV band, and bolometric energy of \eom from the OM observations. In combination with the 18 detected \textit{TESS} flares, we estimate that X-ray flares on TIC~277 can be observed with \textit{XMM-Newton} at least about once a day. %The bolometric and soft-X-ray flare energies are within an order of magnitude within each other, which places them among other stars with simultaneous X-ray and optical flare detections, although, typically, the white light emission dominates over soft X-ray~\citep{guarcello2019simultaneous, kuznetsov2021stellar}.

An indication of an unusual corona stems from its X-ray luminosity relative to bolometric luminosity $L_X/L_{\rm bol}=$\LXLbol, which is about an order of magnitude lower than is typical for saturated activity M dwarfs. TIC 277 is an extreme rapid rotator with a rotation period of only $4.56\,$h. The detected high-latitude flare in \textit{TESS}~\citep{ilin2021giant} may hence be a product of high-latitude flux emergence driven by its fast rotation~\citep{weber2016modeling,weber2017suppression}. This effect is closely related to polar updraft migration, which suppresses coronal emission in equatorial regions~\citep{stepien2001rosat}. However, less than ten flare latitudes on stars other than the Sun are known to date~\citep{wolter2008doppler, ilin2021giant, johnson2021simultaneous}. Localization of more flares on the surfaces of a broader range of stars will allow us to resolve if polar updraft indeed occurs in these stars.

>>>>

\author[]{Desmond Dsouza}
\affiliation{Leibniz Insitute for Astrophysics Potsdam (AIP), An der Sternwarte 16, 14482 Potsdam, Germany}
\affiliation{Institute for Physics and Astronomy, University of Potsdam, Karl-Liebknecht-Strasse 24/25, 14476 Potsdam, Germany}

@@

For the X-ray flare in Fig.~\ref{fig:lightcurves}, we obtain a total energy of \eepic in the $0.2-2\,$keV band \textcolor{red}{by inserting} $L_{X,\rm quiescent}=$\LXquiet\unskip, $L_{X,\rm flaring}=$\LXflaring\unskip, and $\Delta t =$\xraydur\unskip\,(see Fig.~\ref{fig:lightcurves}) in Eq.~\ref{eq:xrayflare}. The energy of the corresponding OM flare is \eom\unskip. We use the 36ks observing baseline of \textit{XMM-Newton} to calculate a flare rate in OM. Fig.~\ref{fig:ffd} illustrates that it is roughly consistent, if slightly higher, than the extension of \textit{TESS}' \textcolor{red}{FFD} to lower energies. About one flare per day \textcolor{red}{with energies} of the order of the observed one or above can therefore be expected in future X-ray observations of TIC~277. \textcolor{red}{However, usually, not all optical flares have X-ray counterparts}.~\citet{paudel2021simultaneous}. \citet{guarcello2019simultaneous, kuznetsov2021stellar} find moderate correlation between flare energies in I data and \textit{XMM-Newton}, but the majority of their sampled flare energies are two orders of magnitude \textcolor{red}{higher than of} the flare in this work, both in X-ray and optical. \citet{kuznetsov2021stellar} compare primary Kepler mission observations with \textit{XMM-Newton} to find a similar result. We note that the X-ray flare light curve follows the typical fast-rise exponential decay shape, and is therefore unlikely to be eclipsed by the star.%~\citep{johnstone2012soft}.

\begin{figure}
    \script{FIGURE_tess_ffd.py}
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/277539431_tess_ffd.png}
        \caption{
         Cumulative flare frequency distribution of \textit{TESS} flares (black dots), with a power law fit (green line), and uncertainties (dotted lines). The rate of OM flares (gray square) is calculated from the observing baseline of \textit{XMM-Newton}. \textcolor{red}{The flare energies are corrected for the pass bands of \textit{TESS} and OM, i.e., given as bolometric flare energies~(see Section~\ref{sec:methods:flareenergies}).} 
        }
        \label{fig:ffd}
    \end{centering}
\end{figure}

@@

Bearing in mind the relatively low two-temperature resolution of TIC~277's X-ray spectrum and the dearth of X-ray spectra for stars at the bottom of the main sequence, it shows a coronal temperature make-up common for mid-to-late M dwarfa in the saturated activity regime\textcolor{red}{, which} it will possibly keep for gigayears until it spins down to very low rotation rates~\citep{medina2022galactic, engle2023living}.

@@

At later spectral type, a decline in coronal activity is expected as the magnetosphere transitions from a stellar to a planetary one~\citep{pineda2017panchromatic}. \citet{stelzer2022first} investigated late M and L dwarfs with eROSITA to show that $L_X/L_{\rm bol}$ increases towards later spectral types~\citep{magaudda2022firsta}, but begins to decrease again from M7 downward. In their compilation, together with previous work from \citet{stelzer2012ultracool, cook2014trends, deluca2020extras, williams2014trends, berger2008simultaneous}, TIC~277 appears at the lower end\textcolor{red}{, but consistent with} the $L_X/L_{\rm bol}$ distribution for M7 dwarfs. However, the rotation periods are unknown for most of these late M dwarfs. Sometimes projected rotation periods are known~\citep{cook2014trends}, but true periods are only detected in a handful of these stars, which we list in Table~\ref{tab:latmlxlbol}. TIC~277 appears \textcolor{red}{at the lower bound of} this small sample. \textcolor{red}{In Magaudda et al. (in prep.), rapidly rotating late M dwarfs observed with eROSITA appear with a range of luminosities centered around $L_X/L_{\rm bol}\approx 10^{-4}$.}

\begin{table*}[]
    \centering
    \caption{Late M dwarfs with known rotation period and X-ray luminosity.}
    \begin{tabular}{lcccc}
        ID & SpT &$P_{\rm rot}$ [d] & $\log_{10} \frac{L_X}{L_{\rm bol}} $ & energy range [keV]\\\hline
        TIC 277 (this work) & M7 & 0.19 & $-4$ & $0.2-2$ \\
        TRAPPIST-1 [1] & M8 & 3.3 & $-4$ & $0.3-10$\\
        LHS 248 [2] & M6.5 & 0.46 & $-3.9$ & $0.2-2$\\
        NLTT 33370 AB [3] & M7 & 0.16 & $-3.5$ & $0.2-2$ \\
        LP 412-31  [4] & M8 & 0.61 & $-3.1$& \textcolor{green}{soft X-ray band?}\\\hline
    \end{tabular}
     \footnotesize
     \vspace{0.1cm}\newline
    [1] \citet{brown2023coronal}, 
    [2] \citet{cook2014trends}, 
    [3] \citet{williams2015simultaneous}, 
    [4] \citet{stelzer2006simultaneous}    
    %3.3 d from \citep{luger2017sevenplanet}
    \label{tab:latmlxlbol}
\end{table*}

@@

\section{Summary and Conclusions}
\label{sec:summary}
We investigated whether TIC 277, a rapidly rotating M7 dwarf with a flare localized at $81^{\circ}$ latitude, shows coronal or flaring properties that could explain the occurrence of this flare so close to the rotational pole. We obtained 36 ks of EPIC and OM observations with \textit{XMM-Newton}, and studied the five Sectors of red-optical 27-day light curves provided by \textit{TESS}. We found a mean quiescent coronal temperature of \Tqmean, and a flaring rate $f(> \log_{10} E_{\rm flare, bol} = 31.5\,\rm{erg})\approx$\ffdr with an energy distribution with a slope of \ffdalpha, all typical of M dwarfs in the saturated regime. We also detected a simultaneous X-ray and optical flare with an energy of \eepic in the $0.2-2\,$keV band, and bolometric energy of \eom from the OM observations. In combination with the 18 detected \textit{TESS} flares, we estimate that X-ray flares on TIC~277 can be observed with \textit{XMM-Newton} at least about once a day. %The bolometric and soft-X-ray flare energies are within an order of magnitude within each other, which places them among other stars with simultaneous X-ray and optical flare detections, although, typically, the white light emission dominates over soft X-ray~\citep{guarcello2019simultaneous, kuznetsov2021stellar}.

An indication of an unusual corona stems from its X-ray luminosity relative to bolometric luminosity $L_X/L_{\rm bol}=$\LXLbol. \textcolor{red}{It is an order of magnitude lower than the canonical value for saturated activity M dwarfs, but representative of late M dwarfs of spectral type M7 and later}. TIC 277 is an extreme rapid rotator with a rotation period of only $4.56\,$h. The detected high-latitude flare in \textit{TESS}~\citep{ilin2021giant} may hence be a product of high-latitude flux emergence driven by its fast rotation~\citep{weber2016modeling,weber2017suppression}. \textcolor{red}{The low X-ray luminosity and the high-latitude flare taken together with the star's fast rotation, saturated flaring activity, and equator-on viewing angle can be explained by polar updraft migration, which suppresses coronal emission in equatorial regions, and confines the corona and flaring regions to near-polar regions~\citep{stepien2001rosat}. }

\textcolor{red}{Overall, }less than ten flare latitudes on stars other than the Sun are known to date~\citep{wolter2008doppler, ilin2021giant, johnson2021simultaneous}. Localization of more flares on the surfaces of a broader range of stars will allow us to resolve if polar updraft migration indeed occurs in these stars.
